---
title: "Case Study 2"
author: "James Greene"
date: "2025-10-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This case study will examine movie ratings from MovieLens users. The data set we will examine the demographics of the reviewers, the genres of the movie titles reviewed, and look for relationships in the data to determine what makes a movie popular. 

```{r packages}
#install packages
packages <- c('tidyverse',
              'ggplot2',
              'stringr',
              'dplyr',
              'magrittr',
              'ggthemes')
install.packages(packages)
```

```{r libraries}
library(tidyverse)
library(ggplot2)
library(stringr)
library(dplyr)
library(magrittr)
library(ggthemes)
```

```{r load data}
movielens = 'https://raw.githubusercontent.com/dnchari/DS501_MovieLens/master/Results/unifiedMLDataMulti.csv'
mlData = read.csv(movielens)
```

First, we'll take a quick look at the overall rating distribution, with a simple histogram plotting the overall count of the frequency of each score.   

```{r overall rating distribution plot}
#plot distribution of ratings
ggplot(mlData,
       aes(x = rating)) + 
  geom_histogram(binwidth = .5, fill = "skyblue", color = "black") +
  labs(title = "Dist. of Movie Rating", 
       x = "Rating",
      y = "Frequency") +
  theme_dark()
```
Here we can see overall, the Movie Lens reviewers have a mild preference for ratings movies a 4 out of 5, but we can't make any specific conjectures, due to the mildly normal distribution of the rating counts. 

We can make some initial comparisons by looking at the overall rating distributions between men and women, and overall rating distribution by occupation and age.

```{r facet wrap gender and occupation}
ggplot(mlData,
       aes(x = rating)) + 
  geom_histogram(binwidth = .5, fill = "skyblue", color = "black") +
  labs(title = "Dist. of Movie Rating by Gender", 
       x = "Rating",
      y = "Frequency") +
    facet_wrap(~gender)
  theme_dark()

ggplot(mlData,
       aes(x = rating)) + 
  geom_histogram(binwidth = .5, fill = "skyblue", color = "black") +
  labs(title = "Dist. of Movie Rating by Occupation", 
       x = "Rating",
      y = "Frequency") +
    facet_wrap(~occupation)
  theme_dark()
```  

Based on these simple visualizations on review frequency, we can make some initial assumptions that our movie ratings will skew towards the preferences of male students. 

##Problem 1: Analysis of the Movie Lens data set

Our first step in our analysis of the Movie Lens data is identifying popularity based on user ratings. This can be done in a handful of ways. 

First, we can simply calculate movies based on their average rating. On this basis, we can make a simple list of total movies with an average rating of 4.5 or higher. 


```{r favorites}
#calculating raw average rating
averageRating <- mlData %>%
  group_by(movie_title) %>%
  summarise(avg_rating = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(avg_rating))

avg_ratingDF <- averageRating %>% filter(avg_rating>=4.50)

knitr::kable(avg_ratingDF)
```

Based on this average rating calculation, we've identified fifteen total movies with an average rating of 4.5 or higher. This approach is a little limited because the computation of  an average score, in general, are easily skewed at small quantities. To help manage this risk, we can filter for films with a threshold of at least 50 reviews. 

```{r}
#fitered average rating to min number of reviews

min_ratings <- 50
filteredavgRating <- mlData %>%
  group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))
```

```{r tables of avg favorites}
top10avg <- head(filteredavgRating, 10)
knitr::kable(top10avg, caption = "Top 10 Movies by Average Rating")
```

We can also examine the median review score, to see if that changes our list of most popualr movies at all. Based on median score, we have an alternative list of top 10 'favorite' films. 

```{r median calcuations}
#median 
medianRating <- mlData %>%
  group_by(movie_title) %>%
  summarise(median_rating = median(rating, na.rm = TRUE)) %>%
  arrange(desc(median_rating))


med_ratingDF <- medianRating %>% filter(median_rating>=4.50)

top_10_median <- head(med_ratingDF, 10)

knitr::kable(top_10_median, caption = "Top 10 movies by Median rating")
```

In this calcuiation, we see Taiwanese new wave title 'Aiqing wansui (1994)' take the top spot. Examining the distribution of the votes for this movie, we see only 1 observation, of 1 reviewer. If only 1 person has reviewed a movie, it's hard for us to conclude it's very popular, since more popular movies will have a larger audience. To adjust for this, let's look at this same median calculations, this time passed through a filtering of a minimum of 50 votes.

```{r plot dist for single title}
aiqingWansui <- mlData %>% filter(movie_title =='Aiqing wansui (1994)')
```






```{r filtered median ratings}
filteredmedianRating <- mlData %>%
  group_by(movie_title) %>%
  summarise(
    medianRating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(medianRating))
```

```{r top 10 median ratings}
top_10_median <- head(filteredmedianRating, 10)
knitr::kable(top_10_median, caption = "Top 10 Movies by Median Rating")
```

Now lets dig into these ratings a bit further, breaking down by gender and age. First, lets look at the highest average movies rated by men. We've noticed that none of the average ratings in the Men's favorite movie list attains an average greater than 4.5 rating, so we'll just look at the top 10 over all, their average rating, and the number of votes the title has in the data. 

```{r}
MensDF <- mlData %>% filter(gender == "M")
favMensDF <- MensDF%>%
 group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))

top_10Mens <- head(favMensDF, 10)

knitr::kable(top_10Mens, caption = "Top 10 Movies amoung Men")
```

Next, we'll repeat the same analysis for the women ratings. 

```{r women's top 10}
WomensDF <- mlData %>% filter(gender == "F")
favWomensDF <- WomensDF%>%
 group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))

top_10Womens <- head(favWomensDF, 10)

knitr::kable(top_10Womens, caption = "Top 10 Movies amoung Women")

```

We can drill down a little deeper in see if there's an age dynamic along with gender, in terms of movie preferences. 

```{r prepare dataframes}
over30men <- MensDF %>% filter(age>30)
over30women <- WomensDF %>% filter(age>30)
under20men <- MensDF %>% filter(age<20)
under20women <-  WomensDF %>% filter(age<20)
```

```{r create tables of 10 ten average scores}
#over30 men
over30MenDF <- over30men%>%
 group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))

over30Mentop10 <- head(over30MenDF, 10)

knitr::kable(over30Mentop10, caption = "Top 10 Movies amoung Men over 30")

#over30women
over30womenDF <- over30women%>%
 group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))

over30Womentop10 <- head(over30womenDF, 10)

knitr::kable(over30Womentop10, caption = "Top 10 Movies amoung Women over 30")

#under20men
under20MenDF <- under20men%>%
 group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))

under20Mentop10 <- head(under20MenDF, 10)

knitr::kable(under20Mentop10, caption = "Top 10 Movies amoung Men under 20")

#under20women
under20womenDF <- under20women%>%
 group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = n(),
    .groups = "drop"  # optional: drops grouping after summarise
  ) %>%
  filter(rating_count >= min_ratings) %>%
  arrange(desc(avg_rating))

under20Womentop10 <- head(under20womenDF, 10)

knitr::kable(under20Womentop10, caption = "Top 10 Movies amoung Women under 20")

```

## Problem 2: Expand our investigation to histogram

```{r histograms}
#histogram of avg_rating distribution
ggplot(averageRating) + 
  geom_histogram(aes(x = avg_rating, binwidth = .5)) +
  labs(title = "Average Ratings of all Movies",
       x = "Average Rating",
       y = "Frequency") + 
  theme_clean()

#histogram of filtered_rating distribution
ggplot(filteredavgRating) + 
  geom_histogram(aes(x = avg_rating, binwidth = .5)) +
  labs(title = "Average Ratings",
       subtitle = "min. 50 votes",
       x = "Average Rating",
       y = "Frequency") +
  theme_clean()
```

```{r median rating}
ggplot(filteredmedianRating) +
  geom_histogram(aes(x=medianRating, binwidth = .5)) +
  labs(title = "Median Rating",
       subtitle = "min. 50 votes",
       x = "Median Rating",
       y= "Frequency") +
  theme_clean()
```

## Problem 3: Correlation: Men versus women

```{r men over 30 histogram}
ggplot(over30MenDF) +
geom_histogram(aes(x=avg_rating, binwidth = .5)) +
  labs(title = "Median Rating",
       subtitle = "Men Over 30",
       x = "Median Rating",
       y = "Frequency") +
  theme_clean()
```

```{r women over 30 histogram}
ggplot(over30womenDF) +
geom_histogram(aes(x=avg_rating, binwidth = .5)) +
  labs(title = "Median Rating",
       subtitle = "Women Over 30",
        x = "Median Rating",
       y = "Frequency") +
  theme_clean()
```

```{r men under 20 histogram}
ggplot(under20MenDF) +
geom_histogram(aes(x=avg_rating, binwidth = .5)) +
  labs(title = "Median Rating",
       subtitle = "Men Under 20",
        x = "Median Rating",
       y = "Frequency") +
  theme_clean()
```

```{r women under 20 histogram}
ggplot(over30MenDF) +
geom_histogram(aes(x=avg_rating, binwidth = .5)) +
  labs(title = "Median Rating",
       subtitle = "Women under 20", 
       x = "Median Rating",
       y = "Frequency") +
  theme_clean()
```


```{r correlation matrix between age and median rating}
```

```{r table of the movies by the number of ratings each movie recieved}
ratingCounts <- mlData %>%
  group_by(movie_title) %>%
  summarise(ratingCount = sum(!is.na(rating)), .groups = "drop") %>%
  arrange(desc(ratingCount))

knitr::kable(head(ratingCounts, 10), caption = "Top 10 Movies by Rating Count")
```

```{r distribution of ratings for all movies}
ggplot(ratingCounts) +
  geom_histogram(aes(x = ratingCount, binwidth = .5)) +
  labs(title = "Rating Counts",
       subtitle = "Distribution of votes per movie", 
       x = "# of Ratings") +
  theme_clean()
```

```{r more ratingCount for top 10 only}
ratingCounts10 <- (head(ratingCounts, 10))

ggplot(ratingCounts10, aes(x = reorder(movie_title, ratingCount), y = ratingCount)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Movies by Rating Count",
    subtitle = "Distribution of votes per movie (most rated)",
    x = "Movie Title",
    y = "# of Ratings"
  ) +
  theme_clean()
```

```{r repeat rating count for movies rated more than 100 times}
min_ratings2 <- 100

filteredratingCounts <- mlData %>%
  group_by(movie_title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    rating_count = sum(!is.na(rating)),
    .groups = "drop"
  ) %>%
  filter(rating_count >= min_ratings2) %>%
  arrange(desc(avg_rating))

knitr::kable(head(filteredratingCounts, 10), caption = "Top 10 Movies by Rating Count, min. 100 votes")
```

```{r}
top25_filtered <- head(filteredratingCounts, 25)
  

ggplot(top25_filtered, aes(x = reorder(movie_title, rating_count), y = rating_count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Filtered Movies by Rating Count",
    subtitle = "min. 100 votes",
    x = "Movie Title",
    y = "# of Ratings"
  ) +
  theme_clean()
```

## Problem 4: Open Ended Question: Business Intelligence 

```{r explore age}


